---
title: "Project proposal"
author: "ARHF"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
## Add any additional packages you are using here
```

## 1. Introduction

The lifestyle dataset was curated by Omar Essa. While the specific method of data collection is not documented, it appears to come from an experimental study that tracked participants’ health and fitness metrics over time. Participants’ physiological and workout-related data were recorded, likely through wearable fitness devices or controlled monitoring during exercise sessions.

Our subset of the dataset includes 11 variables describing demographic, physical, and workout information for each participant. These include:

Age, Gender, Weight, and Height — basic demographic and body measurements.

Max_BPM, Avg_BPM, and Resting_BPM — heart rate data capturing intensity and recovery.

Session_Duration and Calories_Burned — indicators of workout duration and energy expenditure.

Workout_Type — the kind of exercise performed (e.g., Strength, HIIT, Cardio).

BMI — a calculated measure of body fat based on weight and height.

For this research project we are looking to see these questions?
1. How do different workout types (Strength, HIIT, Cardio) affect an individual’s heart rate response (Max_BPM, Avg_BPM, and Resting_BPM)?

2. How do physical characteristics such as BMI, weight, and age influence calories burned and heart rate during workouts?

## 2. Data

There are 54 columns and 2000 rows.

```{r glimpse-data}

Final_data <- read_csv("../data/Final_data.csv")
glimpse(Final_data)
```

## 3. Data analysis plan

Text goes here.
- What variables will you visualize to explore your research questions?
- Will there be any other data that you need to find to help with your research question?
- Very preliminary exploratory data analysis, including some summary statistics and visualizations, along with some explanation on how they help you learn more about your data. (You can add to these later as you work on your project.)
- The data visualization(s) that you believe will be useful in exploring your question(s). (You can update these later as you work on your project.)


```{r}
# select variables we wanna use 
#Code goes here
# Code to calculate summary statistics
# Code for a visualization
```
