---
title: "Project memo"
author: "Aasya, Rachel, Fraol, Hannah"
output: github_document
---

This document should contain a detailed account of the data clean up for your data and the design choices you are making for your plots. For instance you will want to document choices you've made that were intentional for your graphic, e.g. color you've chosen for the plot. Think of this document as a code script someone can follow to reproduce the data cleaning steps and graphics in your handout.

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(hexbin)
```

## Data Clean Up Steps for Overall Data

### Step 1: Read the dataset
We first look at the full lifestyle dataset and checked its structure to understand the variables and types.

```{r read}
Final_data <- read_csv("../data/Final_data.csv")
glimpse(Final_data)
```

### Step 2: Select 11 variables for our research questions
We chose these 11 variables out of 50 to investigate our particular research questions about workout intensity, heart rate, and physical characteristics.

```{r variables}
Final_data |> 
  select(Age, Gender, `Weight (kg)`, `Height (m)` , Max_BPM, Avg_BPM, Resting_BPM, `Session_Duration (hours)` , Calories_Burned, Workout_Type, BMI) |>
  summary()
```

### Step 3: Create BMI groups
We created BMI groups in our data to be able to investigate the overall trends in BMI more clearly for calorie burn. We used standard BMI cutoffs.

```{r BMI-groups}
Final_data <- Final_data %>%
  mutate(BMI_group = case_when(
      BMI < 18.5                     ~ "A.Underweight (<18.5)",
      BMI >= 18.5 & BMI <= 24.9      ~ "B.Normal (18.5–24.9)",
      BMI >= 25   & BMI <= 29.9      ~ "C.Overweight (25.0–29.9)",
      BMI >= 30                      ~ "D.Obese (≥30.0)",
  )
  ) %>%
  filter(!is.na(BMI_group))
```


## Plots

### ggsave example for saving plots

```{r starwars ggsave example}
p1 <- starwars |>
  filter(mass < 1000, 
         species %in% c("Human", "Cerean", "Pau'an", "Droid", "Gungan")) |>
  ggplot() +
  geom_point(aes(x = mass, 
                 y = height, 
                 color = species)) +
  labs(x = "Weight (kg)", 
       y = "Height (m)",
       color = "Species",
       title = "Weight and Height of Select Starwars Species",
       caption = paste("This data comes from the starwars api: https://swapi.py43.com"))


ggsave("example-starwars.png", width = 4, height = 4)

ggsave("example-starwars-wide.png", width = 6, height = 4)
```


#### Final Plot 1: Calories Burned by Workout Type (Violin + Boxplots)

We decided to do a violin plot with a boxplot overlaid investigating how workout type could affect the calories burned of an individual. Incorporating feedback from our presentation, we outlined the yellow color to make it more clear. We used the viridis palette for color-blind viewers.

```{r final-plot1, fig.alt= " Violin and boxplots comparing calories burned with color showing workout type (Cardio, HIIT, Strength, and Yoga). The y axis shows calories burned and the x axis shows workout type. This chart highlights differences in intensity, with HIIT showing the highest values with a longer violin plot showing variation across values and the Yoga being the lowest intensity workout with a narrow violin plot showing less variation across data."}
ggplot(Final_data, aes(x = Workout_Type, y = Calories_Burned, fill = Workout_Type)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15,
               color = "black",     
               fill = "white",  
               outlier.color = "black", 
               outlier.size = 2,alpha = 0.9) +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "Calories Burned by Workout Type",
    subtitle = "Workout types vary in calorie burn, with HIIT the most intense",
    x = "Workout Type",
    y = "Calories Burned",
    fill = "Workout Type"
  ) +
  theme(legend.position = "right")

ggsave("/cloud/project/workouttype.pdf")
ggsave("/cloud/project/workouttype.png")
```

#### Final Plot 2: Weight vs Max BPM by Workout Type, Faceted by Gender

We created a plot showing how weight can affect max BPM, and how it changed across workout type and gender. After getting feedback when we presented we faceted it by gender which makes trends clearer. Workout type is used with a viridis palette for accessibility. We also used loess smoothing curves to depict the trends better than many overlapping points.

```{r final-plot2, fig.alt = "A Faceted smoothed line plot of weight on the x axis and max BPM on the y axis, with separate panels for men and women. Each line represents each workout type. The plot highlights peaks around the middle weights with variation among gender and workout types. Strength and Yoga reach the highest values of BPM for Females, and Strength and HIIT for Males"}

ggplot(Final_data, aes(x = `Weight (kg)`, y = Max_BPM, color = Workout_Type )) +
  geom_smooth(se = FALSE, method = "loess", size = 1.8) +
  scale_color_viridis_d(option = "plasma") +
  labs(
    title = "Weight and Max Heart Rate by Gender",
    subtitle = "Max BPM peaks near mid-weights, with variation across workouts",
    x = "Weight (Kg)",
    y = "Max BPM",
    color = "Workout Type"
  ) +
  facet_wrap(~Gender) +
  theme_minimal(base_size = 13)

ggsave("/cloud/project/maxBPM.pdf")
ggsave("/cloud/project/maxBPM.png")

```

#### Final Plot 3: BMI groups vs. Calories Burned, Facted by Gender

We created a boxplot to compare calories burned across different BMI groups and gender. Since the raw values had wide variation, we used boxplots to show the distribution more clearly. After receiving feedback, we added faceting by gender so the differences between male and female participants are easier to see. Creating BMI groups from the numeric BMI variable also helped make patterns more interpretable when comparing how calorie expenditure differs across categories.

```{r final-plot3, fig.alt = "Boxplots showing calories burned across BMI gourps for male and female participants. The x axes shows the calories burned and the y axes shows the different BMI groups. This chart compare how calorie burn differs across BMI groups and gender, helping correlate physiological characteristics to workout intensity.The plot also shows that both obese groups and underweight groups tend to exhibit more outliers and lower median calorie expenditure compared to mid-range BMI groups"}
ggplot(Final_data, aes(x = Calories_Burned , y = BMI_group, fill = BMI_group)) +
  geom_boxplot() +
  facet_wrap(~ Gender) +
  scale_color_viridis_d(option = "plasma") +
  labs(
    title = "Calories Burned Across BMI Categories by Gender",
    subtitle = "Boxplots reveal how calorie expenditure differs across BMI categories and genders",
    y = "BMI Group",
    x = "Calories Burned",
    fill = "BMI Groups"
  ) +
  theme(legend.position = "right")

ggsave("/cloud/project/BMIgroups.pdf")
ggsave("/cloud/project/BMIgroups.png")
```